Supplementary Materials from: \textit{Innovation and elaboration on the avian tree of life}; Science Advances, in review

### 2023-06-07 Manuscript/main.tex

\subsection{Modeling nested trait covariance}
Addressing the relative contributions of elaboration and innovation to the origins of biodiversity in deep-time requires: 1) large, multivariate datasets to allow exploration of trait covariances at different scales, 2) reliable and efficient computational methods to estimate the major axes of beak shape variation, and 3) a set of mathematical tools that can estimate degrees of elaboration and innovation at any scale.
To meet the first challenge we use an eight-dimensional beak shape morphospace based on a geometric morphometric dataset of 8748 species of birds (described in \cite{hughes2022global}).
To meet the other two challenges we introduce a novel analytical pipeline for measuring elaboration and innovation at the macroevolutionary scale (species within orders) and megaevolutionary scale (orders within the class Aves) (Fig.
 \ref{Fig:cheatsheet} and \ref{Fig:mcmcmcglmm}).
To estimate the major axis of beak shape variation we estimate the evolutionary rate matrix \cite{Houle2017,Machado2020} by fitting Bayesian phylogenetic generalised linear mixed model (pGLMM) models with beak shape as an eight-dimensional response variable \cite{MCMCglmm}.
The major axis can be identified from the leading eigenvector of the posterior distribution of evolutionary rate matrices.
We regard the major axis of the evolutionary rate matrix as a deep-time analogue of the microevolutionary \textbf{G} matrix \cite{steppan2002comparative, robinson2013quantifying} and posit that it represents the among species line of least evolutionary resistance, capturing the effects of historical contingency on multivariate evolution.
This major axis of beak variation can be defined at any phylogenetic scale (i.e. for all species in the phylogeny, or for all species within any clade).
We can then estimate elaboration and innovation based on that major axis: elaboration is defined as the position on the phylogenetic major axis of beak variation and innovation is defined as the distance from the phylogenetic major axis of beak variation (see Methods). %@L_ei_meth_precision.
To assess the roles of elaboration and innovation in beak shape evolution across scales we fit our pGLMMs as nested models in which we define evolutionary rate matrices, and therefore major axes of beak shape variation, for
i) the entire phylogeny of all species included in the data (hereafter the class-wide phylogenetic level),
ii) all species within each of nine clades mapping approximately to super-orders (the super-order level), and
iii) all species within each of 27 clades mapping to taxonomic orders (the order level).
These nested partitions of multivariate trait space provide the basis for all subsequent quantitative estimates of elaboration and innovation across scales.

@@

Phylogenetic distributions of $elaboration_{species}$ and $innovation_{species}$ broadly hold whether comparisons are made at the order, super-order ($elaboration_{species}$ Pagel's $\lambda$ = 0.853; $innovation_{species}$ Pagel's $\lambda$ = 0.914), or class-wide phylogenetic level ($elaboration_{species}$ Pagel's $\lambda$ = 0.935; $innovation_{species}$ Pagel’s $\lambda$ = 0.837; Fig. \ref{Fig:phylogeny_supplement}); \ref{Fig:figure_correlations} %@L_correlation_figure_link)
and in further nested analysis of the order Passeriformes (Fig. \ref{fig_ellipses_passeriformes} and \ref{fig_phylogeny_passeriformes}).
The extent of both elaboration and innovation tends to reduce as we move from class-wide phylogenetic to within-order comparisons.
We regard within-order comparisons as the most informative level of comparison at the macroevolutionary scale
% GT: I added this so that we are clear on our focal analysis. The second part of the sentence is taken directly from one of reviewer 1's comments. TG:TODO: add this to the response letter @@@
because comparisons of species level values derived from the main axis of variation at the level of whole Neornithes might reflect global constraints in beak shape or an axis of innovation among clades (as with the Apodiformes discussed above).
The class-wide phylogenetic major axis of beak variation aligns closely with the raw (non-phylogenetic) major axis in the trait space and primarily describes variation between short, deep, and wide beaks at one extreme, and long, shallow, and narrow beaks at the other.
Our results point to the possibility that within clades the phylogenetic major axes of beak variation follow different directions.

Collectively, our results suggest that elaboration is the most common route by which species beak shape evolves.
While $elaboration_{species}$ is the main mode of beak shape divergence at the macroevolutionary scale %@L_dominant
(Fig. \ref{Fig:phylogeny}), it is also commonly positively correlated with $innovation_{species}$ (Fig. \ref{fable_correlations} \ref{fable_correlations_supplementary}. %@L_correlation_figure_link).
In 14/27 clades we find significant positive correlations with non-significant but positive trends in a further 11 clades.
Only two clades (Falconiformes and Columbiformes) hint at a negative correlation but are clearly non-significant.
The apparent predominance of $elaboration_{species}$ when measured at the macroevolutionary scale is consistent with the idea of evolution along lines of least resistance % @L_genetic_lol
yet the lack of trade-off between $elaboration_{species}$ and $innovation_{species}$ further suggests that evolvability, which is measured by elaboration along a single axis, may in fact be better regarded as multi-dimensional.
Indeed, if divergence can arise simultaneously along multiple axes then this may help to resolve the paradox of apparent shifts in phenotype that have contributed to the diversity of species morphologies across the tree of life.

@@

\subsection{Megaevolution: multiple routes to innovation throughout avian evolutionary history}
If deep-time jumps in beak shape \cite{cooney2017mega} are the result of innovation we would expect to see changes in the orientation of trait space of clades (i.e. orders or super-orders) relative to the phylogenetic major axis of beak variation of the whole class Aves.
To test this, we measured the $elaboration_{clade}$ and $innovation_{clade}$ by translating the phylogenetic major axis of beak variation of each clade's (i.e. super-orders and orders) evolutionary rate matrix onto the class-wide phylogenetic major axis of beak variation so that they shared the same origin in the shapespace.
We then used linear algebra to project %@L_linear_algebra_stuff
a focal clade's phylogenetic major axis of beak variation onto the class-wide phylogenetic major axis of beak variation (Fig.S\ref{Fig:cheatsheet}).
We interpret the measured linear algebraic projection (Euclidean distance along the phylogenetic major axis of beak variation) as the clade's $elaboration_{clade}$ score and the measured linear algebraic rejection (Euclidean distance from the phylogenetic major axis of beak variation) as the clade's $innovation_{clade}$ scor. % @L_euclidean.
A high $innovation_{clade}$ score indicates that the direction of evolution of a clade differs from the direction of its parent clade (e.g. the direction of evolution of an order differs from its parent super-order).
These scores were calculated for each of 4000 posterior pairs of clade vs.
evolutionary rate matrix (Fig. \ref{Fig:ellipses} boxplots).
We found that for most clades the phylogenetic major axis of beak variation is not aligned with the class-wide phylogenetic major axis of beak variation (Fig. \ref{Fig:ellipses}).
Despite the limitations of representing eight-dimensional space in 2D, our plots of the average elliptical representation of the evolutionary rate matrix for the class-wide phylogenetic major axis of beak variation and each super-order and order (Fig. \ref{Fig:ellipses}b) highlight the striking variation in the orientation of beak shape major axes among clades (Fig. \ref{Fig:orthogonality}).
More than half of the assessed clades (4/8 super-orders and 15/27 orders) displayed higher median $innovation_{clade}$ than median $elaboration_{clade}$ scores implying that $innovation_{clade}$ is a more common generator of beak shape diversity than $elaboration_{clade}$ in deep-time (i.e evidence at the clade scale).

@@

Our observations of heterogeneity in the orientation of the phylogenetic major axis of beak variation among clades is further supported by consistent evidence for high amounts of orthogonality of clades relative to the class-wide phylogenetic major axis of beak variation (Fig. \ref{Fig:ellipses} - Fig. \ref{Fig:orthogonality}; note that these comparisons are based on the evolutionary rate matrix). %@L_orientation_phylo
The median angle of the phylogenetic major axis of beak variation for subclades approaches orthogonality, differing from the class-wide phylogenetic major axis of beak variation by $68.14^\circ$ ($95$\% CI: $22.83^\circ$-$89.09^\circ$ - Fig. \ref{Fig:orthogonality}).
Comparisons of orientation among subclades (i.e. orders within super-orders) show similar differences (median = $47.14^\circ$; $95$\% CI: $13.64^\circ$-$87.62^\circ$ - Fig. \ref{Fig:orthogonality}), suggesting that reorientations in trait space are largely unconstrained at the megaevolutionary scale and are no more likely to occur along any one axis than another. 
This differs from previous inference from a subset of our data analysed  without incorporating phylogeny, that implied generally consistent and low dimensionality within clades \cite{cooney2017mega}.
This suggests that there is no megaevolutionary analogue of the genetic line of least resistance and instead that megaevolutionary shifts more likely reflect changes in the adaptive landscape.
Across scales there is remarkable flexibility in the routes to innovation, consistent with the idea that morphological divergence may be less constrained in deep-time than is sometimes assumed \cite{venditti2011multiple}. % @L_occasionaly

@@

We further tested the contributions of elaboration and innovation at different scales by examining the consequences of elaboration and innovation for the observed divergence of bird beak shapes.
Specifically, we calculated the distance to centroid of avian beak morphospace for each species (Fig. \ref{Fig:phylogeny}).
We then used phylogenetic generalised least squares (PGLS; \cite{phylom}) to model distance to centroid as a function of $elaboration_{species}$, $innovation_{species}$ and $elaboration_{clade}$ and $innovation_{clade}$.
We ran the PGLS analysis using both single and multiple predictors.
Our analyses indicate that the majority of variation (r2 = 0.888; Table \ref{Tab:pgls}) in beak shape centroid distance can be explained by a combination of $elaboration_{species}$ and $innovation_{species}$.
When modeled as single predictor models, variation explained is lower for $elaboration_{species}$ than $innovation_{species}$ and $innovation_{species}$ has a steeper slope indicating that although elaboration is overall the dominant model of divergence, innovation leads to greater exploration of morphospace (Table 1).
In contrast, the combination of $elaboration_{clade}$ and $innovation_{clade}$ alone explains only $>$0.01\% of the total variation in beak shape divergence (Table 1).
However, $innovation_{clade}$ is nonetheless an important contributor to total beak shape space.
We reach this conclusion because models with interactions between species and clade metrics provide by far the best model fit overall and removal of terms indicate that the most important interactions are between $innovation_{clade}$  and $innovation_{species}$ followed by $innovation_{clade}$ and $elaboration_{species}$.
The interaction terms show that the relationship between species metrics ($elaboration_{species}$, $innovation_{clade}$) and distance to centroid becomes steeper as $innovation_{clade}$ increases (Table 1 \ref{Tab:pgls})
Overall, the models suggest that although most species evolve via elaboration at the macroevolutionary scale, expansions of beak morphospace are also driven by megaevolutionary scale reorientations of trait space. %@L_PGLS_results

\begin{landscape}
\begin{table}[ht]
\centering
\small
   \begin{tabular}{rrllllllllrr}
     \hline
    Model & $E_{species}$ & $I_{species}$ & $E_{clade}$ & $I_{clade}$ & $E_{species}$ x $I_{clade}$ & $E_{species}$ x $E_{clade}$ & $I_{species}$ x $I_{clade}$ & $I_{species}$ x $E_{clade}$ & $\Delta$AIC & Adj. $r^2$ \\ 
     \hline
   1 & 0.04*** & -0.139*** & 0.002 & -0.002 & 0.064*** & 0.045*** & 0.341*** & 0.093*** & 0.00 & 0.92 \\ 
   2 & 0.257*** & 0.169*** & 0.01 & 0.001 & 0.045*** &  & 0.312*** & 0.034*** & -691.48 & 0.92 \\ 
   3 & 0.108*** & 0.236*** & 0.012 & -0.002 & 0.055*** & 0.034*** & 0.343*** &  & -284.74 & 0.92 \\ 
   4 & 0.254*** & 0.104*** & 0.002 & 0.004 &  & 0.027*** & 0.311*** & 0.052*** & -877.85 & 0.92 \\ 
   5 & 0.1*** & 0.571*** & -0.018* & 0.02*** & 0.054*** & 0.036*** &  & 0.096*** & -1498.36 & 0.91 \\ 
   6 & 0.364*** & 0.973*** & -0.007 & 0.023*** &  &  &  &  & -2230.10 & 0.90 \\ 
   7 & 0.364*** & 0.972*** &  &  &  &  &  &  & -2280.08 & 0.90 \\ 
   8 &  &  & -0.025 & 0.011 &  &  &  &  & -22270.55 & -0.00 \\ 
   9 & 0.286*** &  &  &  &  &  &  &  & -18986.63 & 0.32 \\ 
   10 &  & 0.798*** &  &  &  &  &  &  & -17714.99 & 0.41 \\ 
   11 &  &  & -0.008 &  &  &  &  &  & -22269.34 & -0.00 \\ 
   12 &  &  &  & 0.002 &  &  &  &  & -22269.51 & -0.00 \\ 
      \hline
   \end{tabular}
\caption{Elaboration and innovation as predictors of beak shape distance from centroid.
The table shows parameter estimates of predictors from  12 different phylogenetic generalised least squares (PGLS) models with $\Delta$AIC showing relative model fit following removal of terms. Models 1-5 are multiple regressions including all variables and one or more interaction terms; models 6-8 are are multiple regressions including two or more variables and no interaction terms; models 9-12 are single predictor models. Values in the table are based on PGLS in which Pagel’s $\lambda$ was fixed at 1 (i.e. assuming a Brownian motion model of evolution). Fixing $\lambda$ is necessary to allow model comparison with AIC. We also fitted the same models fixing $\lambda$ to 0.727 (the lowest value of $\lambda$lambda from any individual model) and relative model fit and interpretation of parameters was unaffected. E = elaboration; I = innovation; AIC = Akaike Information Criterion; NS = not significant; - indicates that the term was not included within that model.}
\label{Tab:pgls}
\end{table}
\end{landscape}

@@

\textbf{Beak shape data.} We used the dataset from \cite{cooney2017mega,hughes2022global,chira2020signature} which consists of four landmarks and three curves of 25 semi-landmarks each, taken from 3D scans of the beaks of 8748 bird species.
The landmark data were aligned with Procrustes superimposition using the R package \texttt{Morpho} \cite{Rcore,Morpho} and we removed the size component and controlled for symmetry (see \cite{cooney2017mega,hughes2022global,chira2020signature} for detailed descriptions of data collection and processing).
We ordinated the superimposed landmarks using a principal components analysis (PCA) and selected the eight first dimensions of the resulting PCA.
These eight dimensions contained at least 95\% of variance in each super-order, and 98.7\% of the variance in the whole dataset.
We used a threshold of $>95$\% of variance explained for each super-order because although the overall distribution of the variance on each PC axis in the whole trait-space is decreasing (i.e. $\sigma$ PC1 $>$ $\sigma$ PC2 $>$ $\sigma$ PC3), this is not always true for each super-order.
For example, in the Mirandornithes, PC2, then PC4, and then PC1 contained the most variance (Fig.
\ref{Fig:ellipses} and S\ref{Fig:axes_variance}).
This protocol for selecting the number of PC axes ensures that the resulting trait-space contains enough variance for each super-order (see Fig. \ref{Fig:axes_variance}).
This procedure resulted in an $8748 \times 8$ matrix that contains at least 95\% of the variance in all observed living bird beaks (hereafter, the shapespace). %@L_shapespace_description

@@

\textbf{Elaboration/innovation scores using projection/rejection analyses.} We used the distributions of the 4000 posterior evolutionary rate matrices to run the projection/rejection analyses to obtain elaboration and innovation scores for bird beaks.
We used linear algebra to interpret elaboration and innovation \textit{sensu} \cite{endler2005animal} by using the major axis of the evolutionary rate matrices (referred to throughout the text as the phylogenetic major axis of beak variation) as the ''line of least resistance'', or, specifically here, the line of elaboration for each random term (corresponding to the elaboration axis for each clade).
We projected either species or another phylogenetic major axis of beak variation onto that clade.
We can then interpret where a species or a clade falls on the phylogenetic major axis of beak variation of reference as its $elaboration_{species}$ score, and how far a species is from that phylogenetic major axis of beak variation as its $innovation_{species}$ score.
We calculated the elaboration and innovation scores using two main approaches: 1) by clades (i.e.
to measure the $elaboration_{clade}$/$innovation_{clade}$ at the megaevolutionary scale)where we calculated the projections of the phylogenetic major axis of beak variation of each clade onto the class-wide phylogenetic major axis of beak variation (Fig. \ref{Fig:ellipses}); and 2) by species (i.e.
to measure the $elaboration_{species}$/$innovation_{species}$ at the macroevolutionary scale) where we calculated the projections of each species onto a) the class-wide phylogenetic major axis of beak variation and b) the phylogenetic major axis of beak variation of their parent clade (Fig. \ref{Fig:phylogeny}).
To make the results easier to interpret, we centered and scaled elaboration and innovation scores onto the center of the phylogenetic major axis of beak variation of interest and made the elaboration values absolute.
In this way, we can interpret an elaboration or innovation score within the 0-1 range to be non exceptional, i.e within the 95\% confidence interval of the evolutionary rate matrix.
The full mathematical procedure is described in detail in the supplementary materials (section \ref{supp_projection} - note that the procedure described in the section \ref{supp_projection} is applied to each individual evolutionary rate matrix for each term in the model, e.g. for each 4000 posterior variance matrices individually for each $elaboration_{species}$, $innovation_{species}$, $elaboration_{clade}$ or $innovation_{clade}$ score. %@L_individual_matrices).
Code to reproduce the analyses is available at \href{https://raw.rawgit.net/TGuillerme/dispRity/master/inst/vignettes/Projection_analysis.html}{this \texttt{dispRity} vignette}.

>>>>

\subsection{Modeling nested trait covariance}
Addressing the relative contributions of elaboration and innovation to the origins of biodiversity in deep-time requires: 1) large, multivariate datasets to allow exploration of trait covariances at different scales, 2) reliable and efficient computational methods to estimate the major axes of beak shape variation, and 3) a set of mathematical tools that can estimate degrees of elaboration and innovation at any scale.
To meet the first challenge we use an eight-dimensional beak shape morphospace based on a geometric morphometric dataset of 8748 species of birds (described in \cite{hughes2022global}).
To meet the other two challenges we introduce a novel analytical pipeline for measuring elaboration and innovation at the macroevolutionary scale (species within orders) and megaevolutionary scale (orders within the class Aves; Fig.
 \ref{Fig:cheatsheet} and \ref{Fig:mcmcmcglmm}).
To estimate the major axis of beak shape variation we estimate the evolutionary rate matrix \cite{Houle2017,Machado2020} by fitting Bayesian phylogenetic generalised linear mixed model (pGLMM) models with beak shape as an eight-dimensional response variable \cite{MCMCglmm}.
The major axis can be identified from the leading eigenvector of the posterior distribution of evolutionary rate matrices.
We regard the major axis of the evolutionary rate matrix as a deep-time analogue of the microevolutionary \textbf{G} matrix \cite{steppan2002comparative, robinson2013quantifying} and posit that it represents the among species line of least evolutionary resistance, capturing the effects of historical contingency on multivariate evolution.
This major axis of beak variation can be defined at any phylogenetic scale (i.e. for all species in the phylogeny, or for all species within any clade).
We can then estimate elaboration and innovation based on that major axis: elaboration is defined as the position on the phylogenetic major axis of beak variation and innovation is defined as the distance from the phylogenetic major axis of beak variation (see Methods). %@L_ei_meth_precision.
To assess the roles of elaboration and innovation in beak shape evolution across scales we fit our pGLMMs as nested models in which we define evolutionary rate matrices, and therefore major axes of beak shape variation, for
i) the entire phylogeny of all species included in the data (hereafter the class-wide phylogenetic level),
ii) all species within each of nine clades mapping approximately to super-orders (the super-order level), and
iii) all species within each of 27 clades mapping to taxonomic orders (the order level).
These nested partitions of multivariate trait space provide the basis for all subsequent quantitative estimates of elaboration and innovation across scales.

@@

Phylogenetic distributions of $elaboration_{species}$ and $innovation_{species}$ broadly hold whether comparisons are made at the order, super-order ($elaboration_{species}$ Pagel's $\lambda$ = 0.853; $innovation_{species}$ Pagel's $\lambda$ = 0.914), or class-wide phylogenetic level ($elaboration_{species}$ Pagel's $\lambda$ = 0.935; $innovation_{species}$ Pagel’s $\lambda$ = 0.837; Fig. \ref{Fig:phylogeny_supplement} and \ref{Fig:figure_correlations}) %@L_correlation_figure_link)
and in further nested analysis of the order Passeriformes (Fig. \ref{fig_ellipses_passeriformes} and \ref{fig_phylogeny_passeriformes}).
The extent of both elaboration and innovation tends to reduce as we move from class-wide phylogenetic to within-order comparisons.
We regard within-order comparisons as the most informative level of comparison at the macroevolutionary scale
% GT: I added this so that we are clear on our focal analysis. The second part of the sentence is taken directly from one of reviewer 1's comments. TG:TODO: add this to the response letter @@@
because comparisons of species level values derived from the main axis of variation at the level of whole Neornithes might reflect global constraints in beak shape or an axis of innovation among clades (as with the Apodiformes discussed above).
The class-wide phylogenetic major axis of beak variation aligns closely with the raw (non-phylogenetic) major axis in the trait space and primarily describes variation between short, deep, and wide beaks at one extreme, and long, shallow, and narrow beaks at the other.
Our results point to the possibility that within clades the phylogenetic major axes of beak variation follow different directions.

Collectively, our results suggest that elaboration is the most common route by which species beak shape evolves.
While $elaboration_{species}$ is the main mode of beak shape divergence at the macroevolutionary scale %@L_dominant
(Fig. \ref{Fig:phylogeny}), it is also commonly positively correlated with $innovation_{species}$ (Fig. \ref{fable_correlations} and S8). %@L_correlation_figure_link).
In 14/27 clades we find significant positive correlations with non-significant but positive trends in a further 11 clades.
Only two clades (Falconiformes and Columbiformes) hint at a negative correlation but are clearly non-significant.
The apparent predominance of $elaboration_{species}$ when measured at the macroevolutionary scale is consistent with the idea of evolution along lines of least resistance % @L_genetic_lol
yet the lack of trade-off between $elaboration_{species}$ and $innovation_{species}$ further suggests that evolvability, which is measured by elaboration along a single axis, may in fact be better regarded as multi-dimensional.
Indeed, if divergence can arise simultaneously along multiple axes then this may help to resolve the paradox of apparent shifts in phenotype that have contributed to the diversity of species morphologies across the tree of life.

@@

\subsection{Megaevolution: multiple routes to innovation throughout avian evolutionary history}
If deep-time jumps in beak shape \cite{cooney2017mega} are the result of innovation we would expect to see changes in the orientation of trait space of clades (i.e. orders or super-orders) relative to the phylogenetic major axis of beak variation of the whole class Aves.
To test this, we measured the $elaboration_{clade}$ and $innovation_{clade}$ by translating the phylogenetic major axis of beak variation of each clade's (i.e. super-orders and orders) evolutionary rate matrix onto the class-wide phylogenetic major axis of beak variation so that they shared the same origin in the shapespace.
We then used linear algebra to project %@L_linear_algebra_stuff
a focal clade's phylogenetic major axis of beak variation onto the class-wide phylogenetic major axis of beak variation (Fig. S\ref{Fig:cheatsheet}).
We interpret the measured linear algebraic projection (Euclidean distance along the phylogenetic major axis of beak variation) as the clade's $elaboration_{clade}$ score and the measured linear algebraic rejection (Euclidean distance from the phylogenetic major axis of beak variation) as the clade's $innovation_{clade}$ scor. % @L_euclidean.
A high $innovation_{clade}$ score indicates that the direction of evolution of a clade differs from the direction of its parent clade (e.g. the direction of evolution of an order differs from its parent super-order).
These scores were calculated for each of 4000 posterior pairs of clade vs.
evolutionary rate matrix (Fig. \ref{Fig:ellipses} boxplots).
We found that for most clades the phylogenetic major axis of beak variation is not aligned with the class-wide phylogenetic major axis of beak variation (Fig. \ref{Fig:ellipses}).
Despite the limitations of representing eight-dimensional space in 2D, our plots of the average elliptical representation of the evolutionary rate matrix for the class-wide phylogenetic major axis of beak variation and each super-order and order (Fig. \ref{Fig:ellipses}b) highlight the striking variation in the orientation of beak shape major axes among clades (Fig. \ref{Fig:orthogonality}).
More than half of the assessed clades (4/8 super-orders and 15/27 orders) displayed higher median $innovation_{clade}$ than median $elaboration_{clade}$ scores implying that $innovation_{clade}$ is a more common generator of beak shape diversity than $elaboration_{clade}$ in deep-time (i.e evidence at the clade scale).

@@

Our observations of heterogeneity in the orientation of the phylogenetic major axis of beak variation among clades is further supported by consistent evidence for high amounts of orthogonality of clades relative to the class-wide phylogenetic major axis of beak variation (Fig. \ref{Fig:ellipses}; Fig. \ref{Fig:orthogonality}; note that these comparisons are based on the evolutionary rate matrix). %@L_orientation_phylo
The median angle of the phylogenetic major axis of beak variation for subclades approaches orthogonality, differing from the class-wide phylogenetic major axis of beak variation by $68.14^\circ$ ($95$\% CI: $22.83^\circ$-$89.09^\circ$; Fig. \ref{Fig:orthogonality}).
Comparisons of orientation among subclades (i.e. orders within super-orders) show similar differences (median = $47.14^\circ$; $95$\% CI: $13.64^\circ$-$87.62^\circ$; Fig. \ref{Fig:orthogonality}), suggesting that reorientations in trait space are largely unconstrained at the megaevolutionary scale and are no more likely to occur along any one axis than another. 
This differs from previous inference from a subset of our data analysed  without incorporating phylogeny, that implied generally consistent and low dimensionality within clades \cite{cooney2017mega}.
This suggests that there is no megaevolutionary analogue of the genetic line of least resistance and instead that megaevolutionary shifts more likely reflect changes in the adaptive landscape.
Across scales there is remarkable flexibility in the routes to innovation, consistent with the idea that morphological divergence may be less constrained in deep-time than is sometimes assumed \cite{venditti2011multiple}. % @L_occasionaly

@@

We further tested the contributions of elaboration and innovation at different scales by examining the consequences of elaboration and innovation for the observed divergence of bird beak shapes.
Specifically, we calculated the distance to centroid of avian beak morphospace for each species (Fig. \ref{Fig:phylogeny}).
We then used phylogenetic generalised least squares (PGLS; \cite{phylom}) to model distance to centroid as a function of $elaboration_{species}$, $innovation_{species}$ and $elaboration_{clade}$ and $innovation_{clade}$.
We ran the PGLS analysis using both single and multiple predictors.
Our analyses indicate that the majority of variation (r$^2$ = 0.888; Table \ref{Tab:pgls}) in beak shape centroid distance can be explained by a combination of $elaboration_{species}$ and $innovation_{species}$.
When modeled as single predictor models, variation explained is lower for $elaboration_{species}$ than $innovation_{species}$ and $innovation_{species}$ has a steeper slope indicating that although elaboration is overall the dominant model of divergence, innovation leads to greater exploration of morphospace (Table \ref{Tab:pgls}).
In contrast, the combination of $elaboration_{clade}$ and $innovation_{clade}$ alone explains only $>$0.01\% of the total variation in beak shape divergence (Table \ref{Tab:pgls}).
However, $innovation_{clade}$ is nonetheless an important contributor to total beak shape space.
We reach this conclusion because models with interactions between species and clade metrics provide by far the best model fit overall and removal of terms indicate that the most important interactions are between $innovation_{clade}$  and $innovation_{species}$ followed by $innovation_{clade}$ and $elaboration_{species}$.
The interaction terms show that the relationship between species metrics ($elaboration_{species}$, $innovation_{clade}$) and distance to centroid becomes steeper as $innovation_{clade}$ increases (Table \ref{Tab:pgls}).
Overall, the models suggest that although most species evolve via elaboration at the macroevolutionary scale, expansions of beak morphospace are also driven by megaevolutionary scale reorientations of trait space. %@L_PGLS_results

\begin{landscape}
\begin{table}[ht]
\centering
\small
   \begin{tabular}{rrllllllllrr}
     \hline
    Model & $E_{species}$ & $I_{species}$ & $E_{clade}$ & $I_{clade}$ & $E_{species}$ x $I_{clade}$ & $E_{species}$ x $E_{clade}$ & $I_{species}$ x $I_{clade}$ & $I_{species}$ x $E_{clade}$ & $\Delta$AIC & Adj. $r^2$ \\ 
     \hline
   1 & 0.04*** & -0.139*** & 0.002 & -0.002 & 0.064*** & 0.045*** & 0.341*** & 0.093*** & 0.00 & 0.92 \\ 
   2 & 0.257*** & 0.169*** & 0.01 & 0.001 & 0.045*** &  & 0.312*** & 0.034*** & -691.48 & 0.92 \\ 
   3 & 0.108*** & 0.236*** & 0.012 & -0.002 & 0.055*** & 0.034*** & 0.343*** &  & -284.74 & 0.92 \\ 
   4 & 0.254*** & 0.104*** & 0.002 & 0.004 &  & 0.027*** & 0.311*** & 0.052*** & -877.85 & 0.92 \\ 
   5 & 0.1*** & 0.571*** & -0.018* & 0.02*** & 0.054*** & 0.036*** &  & 0.096*** & -1498.36 & 0.91 \\ 
   6 & 0.364*** & 0.973*** & -0.007 & 0.023*** &  &  &  &  & -2230.10 & 0.90 \\ 
   7 & 0.364*** & 0.972*** &  &  &  &  &  &  & -2280.08 & 0.90 \\ 
   8 &  &  & -0.025 & 0.011 &  &  &  &  & -22270.55 & -0.00 \\ 
   9 & 0.286*** &  &  &  &  &  &  &  & -18986.63 & 0.32 \\ 
   10 &  & 0.798*** &  &  &  &  &  &  & -17714.99 & 0.41 \\ 
   11 &  &  & -0.008 &  &  &  &  &  & -22269.34 & -0.00 \\ 
   12 &  &  &  & 0.002 &  &  &  &  & -22269.51 & -0.00 \\ 
      \hline
   \end{tabular}
\caption{Elaboration and innovation as predictors of beak shape distance from centroid.
The table shows parameter estimates of predictors from 12 different phylogenetic generalised least squares (PGLS) models with $\Delta$AIC showing relative model fit following removal of terms. Models 1-5 are multiple regressions including all variables and one or more interaction terms; models 6-8 are are multiple regressions including two or more variables and no interaction terms; models 9-12 are single predictor models. Values in the table are based on PGLS in which Pagel’s $\lambda$ was fixed at 1 (i.e. assuming a Brownian motion model of evolution). Fixing $\lambda$ is necessary to allow model comparison with AIC. We also fitted the same models fixing $\lambda$ to 0.727 (the lowest value of $\lambda$ from any individual model) and relative model fit and interpretation of parameters was unaffected. E = elaboration; I = innovation; AIC = Akaike Information Criterion; NS = not significant; - indicates that the term was not included within that model.}
\label{Tab:pgls}
\end{table}
\end{landscape}

@@

\textbf{Beak shape data.} We used the dataset from \cite{cooney2017mega,hughes2022global,chira2020signature} which consists of four landmarks and three curves of 25 semi-landmarks each, taken from 3D scans of the beaks of 8748 bird species.
The landmark data were aligned with Procrustes superimposition using the R package \texttt{Morpho} \cite{Rcore,Morpho} and we removed the size component and controlled for symmetry (see \cite{cooney2017mega,hughes2022global,chira2020signature} for detailed descriptions of data collection and processing).
We ordinated the superimposed landmarks using a principal components analysis (PCA) and selected the eight first dimensions of the resulting PCA.
These eight dimensions contained at least 95\% of variance in each super-order, and 98.7\% of the variance in the whole dataset.
We used a threshold of at least 95\% of variance explained for each super-order because although the overall distribution of the variance on each PC axis in the whole trait-space is decreasing (i.e. $\sigma$ PC1 $>$ $\sigma$ PC2 $>$ $\sigma$ PC3), this is not always true for each super-order.
For example, in the Mirandornithes, PC2, then PC4, and then PC1 contained the most variance (Fig.
\ref{Fig:ellipses} and \ref{Fig:axes_variance}).
This protocol for selecting the number of PC axes ensures that the resulting trait-space contains enough variance for each super-order (see Fig. \ref{Fig:axes_variance}).
This procedure resulted in an $8748 \times 8$ matrix that contains at least 95\% of the variance in all observed living bird beaks (hereafter, the shapespace). %@L_shapespace_description

@@

\textbf{Elaboration/innovation scores using projection/rejection analyses.} We used the distributions of the 4000 posterior evolutionary rate matrices to run the projection/rejection analyses to obtain elaboration and innovation scores for bird beaks.
We used linear algebra to interpret elaboration and innovation \textit{sensu} \cite{endler2005animal} by using the major axis of the evolutionary rate matrices (referred to throughout the text as the phylogenetic major axis of beak variation) as the ''line of least resistance'', or, specifically here, the line of elaboration for each random term (corresponding to the elaboration axis for each clade).
We projected either species or another phylogenetic major axis of beak variation onto that clade.
We can then interpret where a species or a clade falls on the phylogenetic major axis of beak variation of reference as its $elaboration_{species}$ score, and how far a species is from that phylogenetic major axis of beak variation as its $innovation_{species}$ score.
We calculated the elaboration and innovation scores using two main approaches: 1) by clades (i.e.
to measure the $elaboration_{clade}$/$innovation_{clade}$ at the megaevolutionary scale)where we calculated the projections of the phylogenetic major axis of beak variation of each clade onto the class-wide phylogenetic major axis of beak variation (Fig. \ref{Fig:ellipses}); and 2) by species (i.e.
to measure the $elaboration_{species}$/$innovation_{species}$ at the macroevolutionary scale) where we calculated the projections of each species onto a) the class-wide phylogenetic major axis of beak variation and b) the phylogenetic major axis of beak variation of their parent clade (Fig. \ref{Fig:phylogeny}).
To make the results easier to interpret, we centered and scaled elaboration and innovation scores onto the center of the phylogenetic major axis of beak variation of interest and made the elaboration values absolute.
In this way, we can interpret an elaboration or innovation score within the 0-1 range to be non exceptional, i.e within the 95\% confidence interval of the evolutionary rate matrix.
The full mathematical procedure is described in detail in the supplementary materials (section \ref{supp_projection};  note that the procedure described is applied to each individual evolutionary rate matrix for each term in the model, e.g. for each 4000 posterior variance matrices individually for each $elaboration_{species}$, $innovation_{species}$, $elaboration_{clade}$ or $innovation_{clade}$ score). %@L_individual_matrices).
Code to reproduce the analyses is available in \href{https://raw.rawgit.net/TGuillerme/dispRity/master/inst/vignettes/Projection_analysis.html}{this \texttt{dispRity} vignette}.

\textbf{Clade orthogonality.} We measured the amount of orthogonality of each random term (i.e.
clade) in our models compared to their parent clade and their parent's parent clade.
For example for Columbiformes, we measured the amount of orthogonality in the evolutionary ratematrices of (i) the Columbimorphae random terms against the phylogeny random terms, (ii) the Columbiformes random terms against the Columbimorphae random terms (parent) and (iii) the Columbiformes random terms the whole phylogeny random terms (parent's parent).
To measure the amount of orthogonality we calculated the angles between the phylogenetic major axis of beak variation of the focal clade and their parent clade for each posterior evolutionary rate matrix and within each clade, by randomly comparing phylogenetic major axes of beak variation within a clade.
We converted each angle measurement into a degree of orthogonality metric where 0 corresponds to flat angles ($0^\circ$ or  $180^\circ$) and 1 to right angles ($90^\circ$ or $270^\circ$).
We then measured the posterior probability of the angles within a clade being different to the angles among clades.
These results are available in Fig. \ref{Fig:orthogonality}.
Note that we did not use the random skewers method here \cite{cheverud2007research} because we were interested in measuring the amount of orthogonality between pairs of evolutionary rate matrices rather than simply measuring whether they were different (but see supplementary materials Fig. S13 for random skewers correlation results). %@L_noskewers
